Predicting the effects of character displacement on food-web dynamics

### 2019-03-22 manuscript_ECD_model.Rmd
added final edits and sent to Seth, JP, and Matt Osmond for friendly reviews.

$^\ast$ Corresponding author; e-mail: matthew.barbour@ieu.uzh.ch, phone: +41 44 635 57 11

@@

*Manuscript type*: Letter. <!-- consider e-Article, depending on length -->

@@

<!-- End title page formatting -->

@@

Ecological character displacement is an adaptive process that generally increases phenotypic diversity. 
Despite the fact that this diversification is due to an eco-evolutionary feedback between consumers competing for shared resources, its consequences for food-web dynamics have not been examined.
Here, I study a model of two consumers competing for two shared resources to examine how character displacement in consumers affects resource abundances and the resilience of food webs to perturbations.
I found that character displacement always strengthened consumer-resource interactions whenever consumers competed for resources that occurred in different habitats.
This increase in interaction strength resulted in lower resource abundances and less resilient food webs.
This occurred under different evolutionary tradeoffs and in both simple and more realistic foraging scenarios.
Taken together, my results show that the adaptive process of character displacement often comes with the ecological cost of decreasing food-web resilience.

@@

Ecological character displacement is an important adaptive process in generating biodiversity [@Schluter2000; @Pfennig2010]. 
This process results from "phenotypic evolution in a species generated or maintained by [exploitative] resource competition with one or more coexisting species” [@Schluter2000]. 
Over the past 40 years, a large body of theoretical [e.g. @Lawlor1976; @Abrams1986; @Doebeli1996; @Taper1985; @McPeek2019] and empirical [reviewed in: @Schluter2000; @Dayan2005; @Stuart2013] work has examined which scenarios lead to phenotypic divergence or convergence of competing consumers.
The general result has been that, if resources are nutritionally substitutable [@Abrams1987; @Fox2008] and there is no other strong source of density dependence acting on consumers [@Abrams1986], then resource competition drives the adaptive divergence of competitors [@Lawlor1976; @Taper1985]. 
This adaptive process is not simply a response to static differences in resource distributions, but creates an eco-evolutionary feedback that drives further differentiation. 
This crucial insight was made by theoretical models that explicitly included resource dynamics as a mediator of competition in driving evolutionary change [@Lawlor1976; @Abrams1986; @Taper1985].

@@

Here, I address this knowledge gap by studying a mathematical model that examines how ecological character displacement affects consumer-resource dynamics in a food-web context. 
I address two questions: (1) How does ecological character displacement affect resource abundances? (2) How does character displacement affect food-web stability? 
To test the generality of these effects, I explored different assumptions of ecological foraging scenarios (Fig. \ref{fig:foraging_scenarios}) and evolutionary trade-offs in consumer foraging traits (Fig. \ref{fig:tradeoff}). 
I found that the adaptive process of character displacement often comes with an ecological cost; resulting in food webs with lower resource availability and less resilient to perturbations.

@@

$$\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2,$$
$$\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2,$$
$$\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1,$$
$$\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2,$$

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ij}$ represents the conversion efficiency of resource $j$ into consumer $i$, and $m_i$ represents the mortality rate of consumer $i$. $F_{ij}(R_j)$ represents consumer $i$'s feeding rate on resource $j$ (i.e. functional response). 
This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example). Importantly, inferences about ecological character displacement can only be made by comparing food webs with and without a competing consumer [@Schluter1992]; therefore, we arbitrarily removed $C_2$ from the system to compare and infer the effects of character displacement on food-web dynamics.

@@

I studied three different foraging scenarios (Fig. \ref{fig:foraging_scenarios}). 
In the first, I assume that consumers can forage for both resources simultaneously and their feeding rates increase linearly with resource abundance, such that:

$$F_{ii}(R_i)=a_{ii}R_i$$

where $a_{ii}$ is the attack rate of consumer $i$ on resource $i$. 
This first scenario is the starting point for many models of resource competition [@MacArthur1972]; however, it does not reflect many food webs where consumers are mobile and their foraging behavior links resources that occur in different habitats [@McCann2005]. 
A simple extension of the above scenario takes the form:

$$F_{ii}(R_i)=w_{ii}a_{ii}R_i$$

where $w_{ii}$ represents the proportion of time consumer $i$ spends foraging in a habitat where only resource $i$ is found (i.e. habitat preference). 
Note that since $w_{ii}$ is a proportion that $w_{ij}=1-w_{ii}$. 
Finally, it is well known that consumer feeding rates often saturate at high resource abundances [@Holling1959; @Rosenzweig1963; @Murdoch2003; @McCann2011] and that consumers do not usually spend a fixed proportion of time in a particular habitat [@McCann2005]. 
To account for these biological realities, I consider a more realistic functional response [derived by @McCann2005]:

$$F_{ii}(R_i,R_j)=\frac{a_{ii}W_{ii}R_i}{1+a_{ii}h_{ii}W_{ii}R_i+a_{ij}h_{ij}W_{ij}R_j}$$

@@

```{r foraging_scenarios, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:foraging_scenarios}**Ecological foraging scenarios.** I examined whether the effect of ecological character displacement on food-web dynamics depended on whether consumers competed for resources that occurred in the same vs. different habitats. In both scenarios, I assumed consumer feeding rate increased linearly with resource density. Later, I relax this assumption and consider a more realistic functional response when resources occurred in different habitats. Note that C~i~ was always pre-adapted to R~i~ (a~ii~ > a~ij~)."}
knitr::include_graphics("ECD_model_foraging_scenarios")
```

@@

Given that character divergence was an expected result, I focus here on its consequences for food-web dynamics across these foraging scenarios.
To do this, I analyzed differences in resource abundances and food-web stability at equilibrium. 
An equilibrium is reached when the four rates of change in the above equations are 0, and solving the system at this point gives equilibrium abundances for each resource ($\hat R_1,\hat R_2$) and each consumer ($\hat C_1,\hat C_2$). 
At the coexistence equilibrium (i.e., where $\hat R_1,\hat R_2, \hat C_1,\hat C_2 > 0$), I calculated resource abundances and performed a local stability analysis [@Otto2007]. 
This stability analysis derives the dominant eigenvalue, $\lambda$, which determines whether (and how readily) the food web will return to equilibrium after a small perturbation.

When possible, I derived analytical expressions for the relationship between consumer attack rates and food-web dynamics. 
To do this, I simplified the model by assuming that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $h=h_{11}=h_{12}=h_{21}=h_{22}$; $m=m_1=m_2$), except that consumer attack rates and their habitat preferences (if present) are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$; $w_{ii}=w_{11}=w_{22}$). 
Controlling for other sources of variability allowed me to isolate the general effects of character divergence.
All mathematical derivations were conducted in @Mathematica and are provided in the Supplementary Information (Appendix S1-3).

To gain insight to the eco-evolutionary feedback generated by character displacement, I conducted simulations using an Adaptive Dynamics approach. 
<!--
The evolutionary dynamics of character displacement have been well characterized in some of the ecological scenarios examined here [@Lawlor1976; @Abrams1986] and are not the focus of this work. 
Still, 
Although divergent character displacement occurs in all scenarios we examined, the magnitude of this divergence may be context dependent.
Therefore, I simulated the eco-evolutionary feedback between consumers and resources using an Adaptive Dynamics approach.
-->
Specifically, at each evolutionary time step, I created a mutant consumer by randomly choosing one and modifying its attack rate on one resource by either subtracting or adding a small constant (0.01 in the following simulations) with equal probability. 
The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. 
This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher than the extremes (when $n>1$, green line in Fig. \ref{fig:tradeoff}) or, conversely, where the two extremes are higher than intermediate investments (when $n<1$, orange line Fig. \ref{fig:tradeoff}). 
When $n=1$, the tradeoff function is linear, and all combinations of $a_{ii}$ and $a_{ij}$ have the same total attack rate (blue line in Fig. \ref{fig:tradeoff}). 
Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. 
If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. 
I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g. $\frac{a_{ii}}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0). 
Unless otherwise noted, I conducted simulations with the following parameter values: *r* = 1; *K* = 4; *e* = 0.8; *m* = 1; *A* = 2; *h* = 0.4; and *w_{ii}* = 0.6. 
I set an initial value of *a~ii~* = 1.2 and *a~ij~* depended on the value of *n*. 
I set initial consumer and resource abundances as: *R~1~* = *R~2~* = 2; *C~1~* = *C~2~* = 1.
All simulations were conducted in R [@R] and the code to reproduce these simulations is publically available on GitHub (https://github.com/mabarbour/ECD_model).

@@

save_plot(filename = "attack_tradeoffs.pdf", plot = tradeoff_plot, 
          base_height = 5, # 6
          base_width = 8) #9
```

```{r tradeoff, fig.pos="H", fig.cap="\\label{fig:tradeoff}**Evolutionary tradeoffs in consumer attack rates.** In each foraging scenario, I explored the effects of three different tradeoffs: intermediate combinations of attack rates (a~ii~, a~ij~) are higher than the extremes (green line, n > 1); extreme combinations of attack rates are higher than intermediate investments (orange line, n < 1); and all combinations of attack rates have the same total attack rate (blue line, n = 1). Points corresponding to attack rates at the beginning of the simulation. Note that C~i~ was always pre-adapted to R~i~ (a~ii~ > a~ij~)."}
knitr::include_graphics("attack_tradeoffs.pdf")
```

@@

In this first scenario, the abundance of resources at equilibrium are equivalent when both consumers and resources are present ($\hat R = \hat R_1 = \hat R_2$), and are determined by the following equation (derived in Appendix S**?**):

$$\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}$$
A key determinant of resource abundance in this model is the consumer's total attack rate, $a_{ii}+a_{ij}$. Therefore, the effect of character displacement on food-web dynamics depends on how the tradeoff influences the evolution of consumer attack rates.

My simulations show that the shape of the tradeoff qualitatively affects the relationship between character displacement and resource abundances (Fig. \ref{fig:plot_fig3}). 
For example, if consumer's are constrained by a linear tradeoff (blue lines), then there is no net change in total attack rate (Fig. \ref{fig:plot_fig3}) and character displacement has no effect on resource abundances (Fig. \ref{fig:plot_fig3}). 
If the tradeoff is concave down (green lines), then resource abundances can actually increase under character displacement (Fig. \ref{fig:plot_fig3}). 
This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{ij}$) and decreases as consumers diverge (Fig. \ref{fig:plot_fig3}). 
When the tradeoff is concave up (orange lines), character displacement suppresses resource abundances due to the increase in total attack rates (Fig. \ref{fig:plot_fig3}). 
Although the equation I derived for resource abundances was for the scenario where both consumers and both resources were present, it accurately predicts the abundance of resources when a single consumer reaches its ESS (triangles on respective colored lines in Fig. \ref{fig:plot_fig3}). This is because a single consumer evolves to be a generalist that has equal attack rates on each resource (triangles at 0.5 along x-axis in Fig. \ref{fig:MacArthur_Plot}), resulting in equivalent resource abundances.

The effect of character displacement on resources corresponds to its impact on food-web stability. 
For example, when character displacement decreases resource abundances (orange points in \ref{fig:plot_fig3}), there is a decrease in food-web stability (\ref{fig:plot_fig3}). 
Similarly, if character displacement does not alter resource abundances (blue and green points in \ref{fig:plot_fig3}), then there is no change in food-web stability (\ref{fig:plot_fig3}).

@@

```{r Stability, include=FALSE, fig.cap="\\label{fig:Stability}Effect of character displacement on local stability when consumers can (A) and cannot (B) forage for both resources simultaneously. Small and large points correspond to initial and evolved strategies, respectively."}
Stability_Mac <- cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + #size=5,
  geom_point(data = filter(MacArthur_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
 #xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

Stability_LS <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  #xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

To examine the ecological consequences of this divergence, I again derive the equilibrium solution when both consumers and both resources are present. 
As before, resource abundances are equivalent at this equilibrium ($\hat R = \hat R_1 = \hat R_2$ when both consumers and resources present), but are now determined by the following equation (derivation in supplementary material):
-->
In this second scenario, I again see that resource abundances are equivalent when both consumers and resources are present ($\hat R = \hat R_1 = \hat R_2$), but are now determined by the following equation (derived in Appendix S**?**):

$$\hat{R}=\frac{1}{w_{ii}a_{ii}+(1-w_{ii})a_{ij}}\cdot\frac{m}{e}$$

>>>>

$^\ast$ Corresponding author; e-mail: matthew.barbour@ieu.uzh.ch<!--, phone: +41 44 635 57 11-->

@@

<!--
\bigskip

*Manuscript type*: Letter.

@@

End title page formatting -->

@@

Ecological character displacement is an adaptive process that generally increases phenotypic diversity. 
Despite the fact that this diversification is due to an eco-evolutionary feedback between consumers competing for shared resources, its consequences for food-web dynamics have not been examined.
Here, I study a model of two consumers competing for two shared resources to examine how character displacement in consumer attack rates affects resource abundances and the resilience of food webs to perturbations.
I found that character displacement always strengthened consumer-resource interactions whenever consumers competed for resources that occurred in different habitats.
This increase in interaction strength resulted in lower resource abundances and less resilient food webs.
This occurred under different evolutionary tradeoffs and in both simple and more realistic foraging scenarios.
Taken together, my results show that the adaptive process of character displacement often comes with the ecological cost of decreasing food-web resilience.

@@

Ecological character displacement is an important adaptive process in generating biodiversity [@Schluter2000; @Pfennig2010]. 
This process is due to "phenotypic evolution in a species generated or maintained by [exploitative] resource competition with one or more coexisting species” [@Schluter2000]. 
Over the past 40 years, a large body of theoretical [e.g. @Lawlor1976; @Abrams1986; @Doebeli1996; @Taper1985; @McPeek2019] and empirical [reviewed in: @Schluter2000; @Dayan2005; @Stuart2013] work has examined which scenarios lead to phenotypic divergence or convergence of competing consumers.
The general result has been that, if resources are nutritionally substitutable [@Abrams1987; @Fox2008] and there is no other strong source of density dependence acting on consumers [@Abrams1986], then resource competition drives the adaptive divergence of competitors [@Lawlor1976; @Taper1985]. 
This adaptive process is not simply a response to static differences in resource distributions, but creates an eco-evolutionary feedback that drives further differentiation. 
This crucial insight was made by theoretical models that explicitly included resource dynamics as a mediator of competition in driving evolutionary change [@Lawlor1976; @Abrams1986; @Taper1985].

@@

Here, I address this knowledge gap by studying a mathematical model that examines how ecological character displacement affects consumer-resource dynamics in a food-web context. 
I address two questions: (1) How does ecological character displacement affect resource abundances? (2) How does character displacement affect food-web stability? 
To test the generality of these effects, I explored different ecological foraging scenarios (fig. \ref{fig:foraging_scenarios}) and evolutionary tradeoffs in consumer attack rates (fig. \ref{fig:tradeoff}). 
I found that the adaptive process of character displacement often comes with an ecological cost; resulting in food webs with lower resource availability and that are less resilient to perturbations.

@@

\begin{equation} \label{eq:1}
  \begin{split}
     & \frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2 \\
     & \frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2 \\
     & \frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1 \\
     & \frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2 \\
  \end{split}
\end{equation}

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ij}$ represents the conversion efficiency of resource $j$ into consumer $i$, and $m_i$ represents the mortality rate of consumer $i$. $F_{ij}(R_j)$ represents consumer $i$'s feeding rate on resource $j$ (i.e. functional response). 
This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates in lakes) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example).

@@

I studied three different foraging scenarios (fig. \ref{fig:foraging_scenarios}). 
In the first, I assume that consumers can forage for both resources simultaneously and their feeding rates increase linearly with resource abundance, such that:

\begin{equation} \label{eq:2}
  F_{ii}(R_i)=a_{ii}R_i
\end{equation}

where $a_{ii}$ is the attack rate of consumer $i$ on resource $i$. 
This first scenario is the starting point for many models of resource competition [@MacArthur1972]; however, it does not reflect many food webs where consumers are mobile and their foraging behavior links resources that occur in different habitats [@McCann2005]. 
A simple extension that accounts for this spatial context takes the form:

\begin{equation} \label{eq:3}
  F_{ii}(R_i)=w_{ii}a_{ii}R_i
\end{equation}

where $w_{ii}$ represents the proportion of time consumer $i$ spends foraging in a habitat where only resource $i$ is found (i.e. habitat preference). 
Note that since $w_{ii}$ is a proportion that $w_{ij}=1-w_{ii}$. 
Finally, it is well known that consumer feeding rates often saturate at high resource abundances [@Holling1959; @Rosenzweig1963; @Murdoch2003; @McCann2011] and that consumers do not usually spend a fixed proportion of time in a particular habitat [@McCann2005]. 
To account for these biological realities, I consider a more realistic functional response [derived by @McCann2005]:     
\begin{equation} \label{eq:4}
  F_{ii}(R_i,R_j)=\frac{a_{ii}W_{ii}R_i}{1+a_{ii}h_{ii}W_{ii}R_i+a_{ij}h_{ij}W_{ij}R_j}
\end{equation}

@@

```{r foraging_scenarios, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:foraging_scenarios}**Ecological foraging scenarios.** I examined whether the effect of ecological character displacement on food-web dynamics depended on whether consumers competed for resources that occurred in the same vs. different habitats. The width of each arrow corresponds to initial attack rates (a~ii~, a~ij~) of consumers. Note that C~i~ was always pre-adapted to R~i~ (a~ii~ > a~ij~). In each scenario, I assumed consumer feeding rate increased linearly with resource abundance. I also relax this assumption and consider a more realistic functional response when resources occurred in different habitats."}
knitr::include_graphics("ECD_model_foraging_scenarios")
```

@@

Given that character divergence occurred across these foraging scenarios, I focus here on its consequences for food-web dynamics.
To do this, I analyzed differences in resource abundances and food-web stability at equilibrium. 
An equilibrium is reached when the rates of change in equation \ref{eq:1} are 0, and solving the system at this point gives equilibrium abundances for each resource ($\hat R_1,\hat R_2$) and each consumer ($\hat C_1,\hat C_2$). 
Importantly, inferences about ecological character displacement can only be made by comparing food webs with and without a competing consumer [@Schluter1992].
Therefore, I compared resource abundances between food webs with ($\hat R_1,\hat R_2, \hat C_1,\hat C_2 > 0$) and without ($\hat R_1,\hat R_2, \hat C_1 > 0$) a competing consumer.
I also compared the local stability of these food webs using standard methods [@Otto2007]. 
This stability analysis derives the dominant eigenvalue, $\lambda$, which determines whether (and how readily) the food web will return to equilibrium after a small perturbation.

When possible, I derived analytical expressions for the relationship between consumer attack rates and food-web dynamics. 
To do this, I simplified the model by assuming that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $h=h_{11}=h_{12}=h_{21}=h_{22}$; $m=m_1=m_2$), except that consumer attack rates and their habitat preferences (if present) are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$; $w_{ii}=w_{11}=w_{22}$). 
Controlling for other sources of variability allowed me to isolate the general effects of character divergence.
All mathematical derivations were conducted in Mathematica [@Mathematica] and are provided in the Supplementary Information (Appendix S1-3).

To gain insight to the eco-evolutionary feedback generated by character displacement, I conducted simulations using an Adaptive Dynamics approach. 
<!--
The evolutionary dynamics of character displacement have been well characterized in some of the ecological scenarios examined here [@Lawlor1976; @Abrams1986] and are not the focus of this work. 
Still, 
Although divergent character displacement occurs in all scenarios we examined, the magnitude of this divergence may be context dependent.
Therefore, I simulated the eco-evolutionary feedback between consumers and resources using an Adaptive Dynamics approach.
-->
Specifically, at each evolutionary time step, I created a mutant consumer by randomly choosing one and modifying its attack rate on one resource by either subtracting or adding a small constant (0.01 in the following simulations) with equal probability. 
The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. 
This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher than the extremes (when $n>1$, green line in fig. \ref{fig:tradeoff}) or, conversely, where the two extremes are higher than intermediate investments (when $n<1$, orange line fig. \ref{fig:tradeoff}). 
When $n=1$, the tradeoff function is linear, and all combinations of $a_{ii}$ and $a_{ij}$ have the same total attack rate (blue line in fig. \ref{fig:tradeoff}). 
Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. 
If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. 
I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g. $\frac{a_{ii}}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0). 
Unless otherwise noted, I conducted simulations with the following parameter values: *r* = 1; *K* = 4; *e* = 0.8; *m* = 1; *A* = 2; *h* = 0.4; and *w~ii~* = 0.6. 
I set an initial value of *a~ii~* = 1.2 and *a~ij~* depended on the value of *n*. 
I set initial consumer and resource abundances to: *R~1~* = *R~2~* = 2; *C~1~* = *C~2~* = 1.
All simulations were conducted in R [@R] and the code to reproduce these simulations is publically available on GitHub (https://github.com/mabarbour/ECD_model).

@@

save_plot(filename = "attack_tradeoffs.pdf", plot = tradeoff_plot, 
          base_height = 5, # 6
          base_width = 7) #9
```

```{r tradeoff, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:tradeoff}**Evolutionary tradeoffs in consumer attack rates.** In each foraging scenario, I explored the effects of three different tradeoffs: intermediate combinations of attack rates (a~ii~, a~ij~) are higher than the extremes (green line, n > 1); extreme combinations of attack rates are higher than intermediate investments (orange line, n < 1); and all combinations of attack rates have the same total attack rate (blue line, n = 1). Points corresponding to attack rates at the beginning of the simulation. Note that C~i~ was always pre-adapted to R~i~ (a~ii~ > a~ij~)."}
knitr::include_graphics("attack_tradeoffs.pdf")
```

@@

In this first scenario (equation \ref{eq:2}), the abundance of resources at equilibrium are equivalent when both consumers and resources are present ($\hat R = \hat R_1 = \hat R_2$), and are determined by the following equation (derived in Appendix S2):

\begin{equation}
  \hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}
\end{equation}

A key determinant of resource abundance in this model is the consumer's total attack rate, $a_{ii}+a_{ij}$. Therefore, the effect of character displacement on food-web dynamics depends on how the tradeoff influences the evolution of consumer attack rates.

My simulations show that the shape of the tradeoff qualitatively affects the relationship between character displacement and resource abundances (fig. \ref{fig:plot_fig3}). 
For example, if consumer's are constrained by a linear tradeoff (blue lines), then there is no net change in total attack rate (fig. \ref{fig:plot_fig3}) and character displacement has no effect on resource abundances (fig. \ref{fig:plot_fig3}). 
If the tradeoff is concave down (green lines), then resource abundances can actually increase under character displacement (fig. \ref{fig:plot_fig3}). 
This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{ij}$) and decreases as consumers diverge (fig. \ref{fig:plot_fig3}). 
When the tradeoff is concave up (orange lines), character displacement suppresses resource abundances due to the increase in total attack rates (fig. \ref{fig:plot_fig3}). 
Although the equation I derived for resource abundances was for the scenario where both consumers and both resources were present, it accurately predicts the abundance of resources when a single consumer reaches its ESS (triangles on respective colored lines in fig. \ref{fig:plot_fig3}). 
This is because a single consumer evolves to be a generalist that has equal attack rates on each resource (triangles at 0.5 along x-axis in fig. \ref{fig:plot_fig3}), resulting in equivalent resource abundances.

The effect of character displacement on resources corresponds to its impact on food-web stability. 
For example, when character displacement decreases resource abundances (orange points in fig. \ref{fig:plot_fig3}), there is a decrease in food-web stability (fig. \ref{fig:plot_fig3}). 
Character divergence may not affect or even increase food-web stability (blue and green lines in fig. \ref{fig:plot_fig3}); however, evolution does not favor strong divergence in these scenarios (blue and green points in fig. \ref{fig:plot_fig3}), which dampens these contingent effects.

@@

```{r Stability, include=FALSE, fig.cap="\\label{fig:Stability}Effect of character displacement on local stability when consumers can (A) and cannot (B) forage for both resources simultaneously. Small and large points correspond to initial and evolved strategies, respectively."}
Stability_Mac <- cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + #size=5,
  geom_point(data = filter(MacArthur_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
 #xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  #geom_hline(yintercept=0, color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

Stability_LS <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  #xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  #geom_hline(yintercept=0, color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

To examine the ecological consequences of this divergence, I again derive the equilibrium solution when both consumers and both resources are present. 
As before, resource abundances are equivalent at this equilibrium ($\hat R = \hat R_1 = \hat R_2$ when both consumers and resources present), but are now determined by the following equation (derivation in supplementary material):
-->
In the second foraging scenario (equation \ref{eq:3}), I again see that resource abundances are equivalent when both consumers and resources are present ($\hat R = \hat R_1 = \hat R_2$), but are now determined by the following equation (derived in Appendix S3):

\begin{equation}
  \hat{R}=\frac{1}{w_{ii}a_{ii}+(1-w_{ii})a_{ij}}\cdot\frac{m}{e}
\end{equation}

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w_{ii}>0.5$ and $a_{ii}>a_{ij}$), then the effective attack rate of consumers ($wa_{ii}+(1-w_{ii})a_{ij}$) will always increase, regardless of the tradeoff (fig. \ref{fig:plot_fig3}). 
Thus, character divergence always results in resource suppression (fig. \ref{fig:plot_fig3}). 
Note that the shape of the tradeoff can modify the effect of character displacement. 
This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached an evolutionary stable strategy (ESS, triangles in fig. \ref{fig:plot_fig3}). 
In contrast, resource abundances reach a similar value when consumers evolve in the presence of a competitor (circles in fig. \ref{fig:plot_fig3}), because character displacement tends to reach a constraint of complete specialization. 
It is worth noting that resource abundances are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in fig. \ref{fig:plot_fig3}). 
This is because consumers actually evolve to be slightly specialized on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in fig. \ref{fig:plot_fig3}).

As seen previously, the effect of character displacement on resource abundances qualitatively corresponds to its effect on food-web stability (fig. \ref{fig:plot_fig3}). 
Specifically, character divergence decreases food-web stability, regardless of the tradeoff in attack rates.
This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (fig. \ref{fig:plot_fig3}). 
For example, when the tradeoff is concave up (orange), the initial two-consumer food web (small circle) is more stable than when there is only one consumer (small triangle); however, this pattern switches by the end of the eco-evolutionary simulation (large points).
